Chess collusion with hyper3 objects

### 2022-08-29 inst/collusion3.Rmd
profile likelihood
defect of the model

@@

```{r,label=H2H3}
H2 <- hyper2()
H3 <- hyper3()
f <- function(lambda,D){
  for(i in seq_len(nrow(cur_matches))){
    white_player <- cur_matches[i,1]
    black_player <- cur_matches[i,2]
    result <- cur_matches[i,3]
    if(result == "1-0"){ # white player wins
       num <- lambda 
       names(num) <- white_player
    } else if(result == "0-1"){  # black player wins
       num <- 1
       names(num) <- black_player
    } else if (result == "1/2-1/2"){ # draw, duh
      num <- c(D,D)
      names(num) <- c(white_player,black_player)
    } else {
      stop("this cannot happen")
    }

@@

>>>>

# Short story

With this model, whatever the value of $D$ and $\lambda$, there is a
pathology which BT cannot deal with.  Still not quite 100% sure but it
seems that the maximum likelihood estimate of the players' strengths
is zero except for Fisher and Petrosian; and indeed we see Fisher at
57% and Petrosian at 42% [with $\lambda = 1.1, D=1.888$].  This on the
grounds that these players never lost playing black (there are some
$2\times 3$ tables at the end).  This might be considered a defect of
the probability model.

# Longer story

@@

```{r,label=H2H3}
f <- function(lambda,D){
  H3 <- hyper3()
  for(i in seq_len(nrow(cur_matches))){
    white_player <- cur_matches[i,1]
    black_player <- cur_matches[i,2]
    result <- cur_matches[i,3]
    if(result == "1-0"){ # white player wins
       num <- lambda 
       names(num) <- white_player
    } else if(result == "0-1"){  # black player wins
       num <- 1
       names(num) <- black_player
    } else if (result == "1/2-1/2"){ # draw, duh
      num <- c(D,D)
      names(num) <- c(white_player,black_player)
    } else {
      stop("this cannot happen")
    }

@@

```{r,testH3,cache=TRUE}
equalp.test(H3,n=1)
```

```{r,comparelikelihoods,cache=TRUE}
maxp(f(1.1,2.1),n=1,give=TRUE)
maxp(f(1.1,2.3),n=1,give=TRUE)
```

A break here.

```{r,label=defineprofilelike,cache=FALSE}
profl <- function(v){
    lambda <- v[1]
    D <- v[2]
    H <- f(lambda,v)
    return(loglik(maxp(H,n=1),H))
}
```

Now apply `profl()`:

```{r,calcproflike,cache=TRUE}
profl(c(1.1,2.1))
#profl(c(1.2,2.1))
```

```{r,calcmanyproflike,cache=TRUE} 
np <- 4
lambda <- seq(from=0.1,to=1.5,len=np)
D <- seq(from=0.1,to=0.4,len=np)
jj <- as.matrix(expand.grid(lambda,D))
LL <- apply(jj,1,profl)
LL
```

```{r,plotll}
LL <- LL - max(LL)
m <- matrix(LL,np,np)
m
filled.contour(lambda,D,m,xlab="lambda",ylab="D",nlevels=40)
```

```{r,plotagain}
res <- cur_matches$V3
res
table(res)
white_wins <- cur_matches[res=="1-0"    ,]
black_wins <- cur_matches[res=="0-1"    ,]
draws      <- cur_matches[res=="1/2-1/2",]
(players <- unique(sort(c(cur_matches[,1],cur_matches[,2]))))
jj <- rep(0,length(players))
names(jj) <- players
jj
plays_white_wins <- jj
plays_white_lose <- jj
plays_white_draw <- jj
plays_black_wins <- jj
plays_black_lose <- jj
plays_black_draw <- jj
for(i in seq_len(nrow(cur_matches))){
  white_player <- cur_matches[i,1]
  black_player <- cur_matches[i,2]
  result       <- cur_matches[i,3]

  if(result == "1-0"){
      plays_white_wins[white_player] %<>% inc
      plays_black_lose[black_player] %<>% inc
  } else if(result == "0-1"){
      plays_white_lose[white_player] %<>% inc
      plays_black_wins[black_player] %<>% inc
  } else if(result == "1/2-1/2"){
      plays_white_draw[white_player] %<>% inc
      plays_black_draw[black_player] %<>% inc
  }
}

plays_white_wins
plays_white_lose
plays_black_wins
plays_black_lose
```

Now we need some way to describe each player's wins/losses playing
white and black:

```{r}
f <- function(player){
out <- matrix(c(
  plays_white_wins[player], plays_white_lose[player], plays_white_draw[player],
  plays_black_wins[player], plays_black_lose[player], plays_black_draw[player]
  ),byrow=TRUE,2,3)
  dimnames(out) <- list(colour=c("white","black"),match=c("win","lose","draw"))
  return(out)
}
f("Fischer")
f("Korchnoi")
f("Petrosian")
M <- f("Aaron")*0
for(p in players){ M <- M + f(p) }
M
```
